---
title: "Population Structure"
author: "John D. Robinson"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    toc: true
linkcolor: blue
---

# Characterizing Genetic Structure across Georgia's Black Bear Populations
In this markdown we'll use some files created in `diversity` to characterize population differentiation across Georgia. Two primary analyses are presented below:

1. Calculation of pairwise $F_{ST}$ 
2. Clustering analyses (PCA and `admixture`)

But first, to avoid including identical genotypes in the pairwise $F_{ST}$ calculations and the `admixture` analyses below, we will first drop the samples that were duplicated in `combined.vcf` - the "survivors", bears that were detected in both 2012 and 2023. We wanted to include those samples in both populations for diversity analyses, but here we'll only keep them in the 2012 group.  

## Deal with "survivors"
```{r checkRepeats}
CGA12 <- read.table("CGA12keep.txt", header = FALSE)
dim(CGA12)
CGA23 <- read.table("CGA23keep.txt", header = FALSE)
dim(CGA23)

CGA23nodup <- CGA23$V1[-which(CGA23$V1 %in% CGA12$V1)]
length(CGA23nodup)

write.table(CGA23nodup, file = "CGA23nodup.txt", 
            quote = FALSE, 
            row.names = FALSE,
            col.names = FALSE)
```

Now extract the 2023 samples only, this way we don't have duplicated genotypes in our analysis.  
```{r engine="bash", comment=""}
#Output the CGA23 only samples
vcftools --vcf ../snpfiltering/PopGen.vcf --keep CGA23nodup.txt \
          --recode --out CGA23nodup

#Grab header and SNP rows
grep -v ^## CGA23nodup.recode.vcf | cut -f 10- > CGA23nodupsnps

#Paste data columns together (using objects made in diversity)
paste CGA12snps CGA23nodupsnps NGAsnps SGAsnps > tempstruct

#Grab header from one of the files
grep ^## CGA12.recode.vcf > structure.vcf
cat tempstruct >> structure.vcf

grep -v ^## structure.vcf | head -1 | tr '\t' '\n' | grep ^UGA > structIID.txt
```

## Pairwise $F_{ST}$ in `PLINK`
For this analysis, we first need to make a population "phenotype" file for `PLINK2`...    
```{r plinkFst}
#Load the metadata file
samps <- read.csv("../input/sampleMetaData_cloneID.csv", header = TRUE)

#Use the IID object created above to build plinkPops
IID <- read.table("structIID.txt", header = FALSE)
FID <- c()
for(i in 1:nrow(IID)) {
  if(strsplit(IID[i,1],split="_")[[1]][1] == "UGAh") {
    FID[i] <- samps$region[which(samps$vcfID == gsub("h","",IID[i,1]))]
  } else {
    FID[i] <- samps$region[which(samps$vcfID == IID[i,1])]
  }

  if(FID[i] %in% c("CGA", "CGA-H")) {
    if(strsplit(IID[i,1],split="_")[[1]][1] == "UGAh") {
      FID[i] <- "CGA12"
    } else {
      FID[i] <- "CGA23"
    }
  }
}
table(FID)
plinkPops <- data.frame(FID=FID, IID=IID, pop=FID)

#Write this as a phenotype file
write.table(plinkPops, file = "plinkPops.txt",
            quote = FALSE,
            row.names = FALSE,
            col.names = FALSE)
system("echo '#FID IID pop' > plinkSampInfo.txt")
system("cat plinkPops.txt >> plinkSampInfo.txt")
```

Now use the `--fst` argument in `PLINK2` to calculate two measures of $F_{ST}$:

1. [Weir & Cockerham (1984)](https://doi.org/10.2307/2408641) - `'method=wc'`
2. The Hudson estimator ([Hudson et al. 1992](https://doi.org/10.1093/genetics/132.2.583), [Bhatia et al. 2013](https://doi.org/10.1101/gr.154831.113)) - `'method=hudson'`  


```{r engine="bash", comment = ""}
plink2 --vcf structure.vcf --fam plinkSampInfo.txt \
        --fst pop 'method=wc' --out WC --silent
        
plink2 --vcf structure.vcf --fam plinkSampInfo.txt \
        --fst pop 'method=hudson' --out HUDSON --silent
```

Read the output into R...  
```{r FstOutput}
read.table("WC.fst.summary", header = FALSE)
read.table("HUDSON.fst.summary", header = FALSE)
```

We see more genetic differentiation between central GA and north / south GA than between north and south GA, despite those two populations being described as separate subspecies (*Ursus americanus americanus* in the north and *U. a. floridanus* in the south). This pattern matches that reported with microsatellite loci in [Hooker et al. (2019)](https://doi.org/10.2192/URSUS-D-18-00025.1).  


## Clustering analyses

### Model-free clustering (PCA)
First using the `adegenet::glPca()` function to visualize clustering across the first three principal components.  
```{r PCA}
library(vcfR)
library(adegenet)
library(scales)

vcf <- read.vcfR("structure.vcf", verbose=TRUE)
my_genind <- vcfR2genind(vcf, pop=plinkPops$pop)

my.pca <- glPca(as.genlight(my_genind@tab), nf = 3)

transparent.cols <- c(alpha("darkgoldenrod1", 0.65),
                      alpha("lightgreen", 0.65),
                      alpha("darkgreen", 0.65),
                      alpha("lightblue", 0.65))
ptcolors <- rep("black", length(my_genind@pop))
ptcolors[my_genind@pop == "NGA"] <- transparent.cols[1]
ptcolors[my_genind@pop == "CGA12"] <- transparent.cols[2]
ptcolors[my_genind@pop == "CGA23"] <- transparent.cols[3]
ptcolors[my_genind@pop == "SGA"] <- transparent.cols[4]

table(ptcolors)

#PC1vPC2
plot(my.pca$scores[,1], my.pca$scores[,2], pch = 19, col = ptcolors,
     xlab = "PC1", ylab = "PC2")
legend("topleft", bty = "n", pch = c(19,19,19), 
       col=transparent.cols,
       legend = c("NGA","CGA-2012","CGA-2023","SGA"), cex = 0.75)

#PC1vPC3
plot(my.pca$scores[,1], my.pca$scores[,3], pch = 19, col = ptcolors,
     xlab = "PC1", ylab = "PC3")
legend("bottomright", bty = "n", pch = c(19,19,19), 
       col=transparent.cols,
       legend = c("NGA","CGA-2012","CGA-2023","SGA"), cex = 0.75)

#PC2vPC3
plot(my.pca$scores[,2], my.pca$scores[,3], pch = 19, col = ptcolors,
     xlab = "PC2", ylab = "PC3")
legend("topleft", bty = "n", pch = c(19,19,19), 
       col=transparent.cols,
       legend = c("NGA","CGA-2012","CGA-2023","SGA"), cex = 0.75)
```

There are four samples from CGA that are clustering with samples from NGA...  
```{r checkCGA}
#Identify CGA samples clustering with NGA...
Qsamps <- which(my.pca$scores[,1] > 0 & ptcolors == transparent.cols[3])
samps[which(samps$vcfID %in% names(Qsamps)),]
```

These samples are actually offspring from a NGA sow that was relocated to CGA. She moved north to Hancock / Greene Co. and mated with a male in the area. These samples were either collected from the cubs in the field during a den check (`73-B` and `74-B`), or collected from bears that were hit by motorists (`83-RK` and `89-RK`). It makes sense that they're clustering with NGA bears, but they are not natural migrants from NGA.  

### Model-based clustering (`admixture`)
Next we'll use the model-based clustering analysis in `admixture` ([Alexander et al. 2009](https://doi.org/10.1101/gr.094052.109)) to infer group membership. This is the same underlying model used by `STRUCTURE`, but the maximum likelihood implementation is much faster.  

We can use [PLINK v. 1.9](https://www.cog-genomics.org/plink/) to create a `.bed` file input for `admixture`, still working from the `structure.vcf` generated above.  
```{zsh recode12}
plink --vcf structure.vcf --recode12 --out ADMIX \
        --double-id --allow-extra-chr --silent
```

We can download `admixture` binaries (for Mac and Linux) from [David Alexander's GitHub site](https://dalexander.github.io/admixture/).  

Now run `admixture` with K=1-6...  
```{r engine="bash", comment =""}
ADMIX="/home/jdrobins/admixture_linux-1.3.0/admixture"

for K in 1 2 3 4 5 6
  do $ADMIX --cv=10 ADMIX.ped $K | tee log${K}.out
done

#Best K choice will give us lowest CV error
grep -h CV log*.out
```

Now let's plot the results...  
```{r pickK}
## First the CV error plot
CVerr <- system("grep -h CV log*.out | cut -f 2 -d :", intern = TRUE)
CVerr <- as.numeric(CVerr)
plot(x = c(1:6), CVerr, type = "b", xlab = "K", ylab = "CV Error")

bestK <- which(CVerr == min(CVerr))
```

The best $K$ appears to be $K =$ `r bestK`, based on our cross-validations. Now let's visualize the results at the different $K$ values...   
```{r ADMIXres}
## Now for admixture proportions
#First read in the sample IDs
adsamps <- read.table("ADMIX.nosex", header = FALSE)
popmap <- data.frame(vcfID = adsamps[,1], pop=NA, sortord=NA)
for(i in 1:nrow(popmap)) {
  if(strsplit(popmap$vcfID[i],split="_")[[1]][1] == "UGAh") {
    popmap$pop[i] <- "CGA12"
  } else {
    popmap$pop[i] <- samps$region[which(samps$vcfID == popmap$vcfID[i])]
    if(popmap$pop[i] %in% c("CGA","CGA-H")) {
      popmap$pop[i] <- "CGA23"
    } 
  }
}

#Plot results with K values from 2 to 6
sortedQ <- colscheme <- vector("list", 6)
for(k in 2:6) {
  bestKQ <- read.table(paste0("ADMIX.",k,".Q"), header=FALSE)
  dim(bestKQ)
  
  
  #Order the color scheme so it lines up with the populations...
  popXcol <- data.frame(pop = c("NGA","CGA23","CGA12","SGA"),
                        col = c("darkgoldenrod1","darkgreen",
                                "lightgreen","lightblue"))
  colscheme[[k]] <- rep(NA, k)
  p <- 1
  while(sum(is.na(colscheme[[k]])) > 0) {
    if(p > 4) {
      if(sum(is.na(colscheme[[k]])) == 1) {
        colscheme[[k]][is.na(colscheme[[k]])] <- "lightgrey"
      } else if(sum(is.na(colscheme[[k]])) == 2) {
        colscheme[[k]][is.na(colscheme[[k]])] <- c("lightgrey","darkgrey")
      } else if(sum(is.na(colscheme[[k]])) == 3) {
        colscheme[[k]][is.na(colscheme[[k]])] <- c("lightgrey","darkgrey","black")
      }
    } else {
      tmp <- which(apply(bestKQ[which(popmap$pop == popXcol$pop[p]),],2,mean) > 0.5)
      if(length(tmp) > 0 & is.na(colscheme[[k]][tmp])) {
        colscheme[[k]][tmp] <- popXcol$col[p]
      }
      rm(tmp)
    }
    p <- p+1
  }
  
  #Set the order for plotting along the x-axis (individuals)...
  popmap$sortord[which(popmap$pop == "NGA")] <- 1
  popmap$sortord[which(popmap$pop == "CGA12")] <- 2
  popmap$sortord[which(popmap$pop == "CGA23")] <- 3
  popmap$sortord[which(popmap$pop == "SGA")] <- 4
  
  sum(is.na(popmap$sortord))
  
  #Sort the samples by population - NGA, CGA12, CGA23, SGA...
  sortedmap <- popmap[order(popmap$sortord),]
  sortedQ[[k]] <- bestKQ[order(popmap$sortord),]
  
  fillcols <- popXcol$col[match(colscheme[[k]], popXcol$col)]
  fillleg <- popXcol$pop[match(colscheme[[k]], popXcol$col)]
  fillcols <- fillcols[!is.na(fillcols)]
  fillleg <- fillleg[!is.na(fillleg)]
  
  barplot(t(as.matrix(sortedQ[[k]])), 
          col=colscheme[[k]], ylab = "Ancestry", xlab = "Individual",
          border=NA, names=sortedmap$pop, cex.names = 0.75)
  legend("topright", bty="n", title = "Genetic Cluster",
         ncol = length(fillcols), 
         fill = fillcols, legend = fillleg, 
         xpd = TRUE, inset = c(0, -0.2))
}
```


### North GA bears in Central GA?
There are four bears in the central GA sample have north GA ancestry (four golden bars in the predominantly green sample set). Let's take a look at these bears...  
```{r NGAinCGA}
#These show up with K = 2
colscheme[[2]]

nga.col <- which(colscheme[[2]] == popXcol$col[which(popXcol$pop == "NGA")])
sortedmap[which(sortedQ[[2]][,nga.col] > 0.5 & 
                  !sortedmap$pop %in% c("NGA","SGA")),]
samps[which(samps$vcfID %in% 
              sortedmap$vcfID[which(sortedQ[[2]][,nga.col] > 0.5 &
                                      !sortedmap$pop %in% c("SGA","NGA"))]),]
```


Those are the same four bears we saw in the PCA. Offspring of a translocated NGA sow that mated with a male bear in the vicinity of Greene and Hancock counties (north of CGA).  



---

## Summary
Overall, the `admixture` analysis supports a solution with $K=3$. The clustering at this $K$ cleanly splits the three populations into separate clusters. Four bears with NGA ancestry are present in our CGA sample, but three of the four are offspring from a NGA sow that was translocted to CGA. The male she mated with in Hancock or Greene County (north of the CGA study region) must also have been of NGA origin (given the lack of evidence for CGA contributions to these samples). All three of these cubs were subsequently killed on the road by motorists, although one cub did make it back to CGA and appeared once in our SCR dataset. The other bear of NGA ancestry was also hit on the road near Augusta (again outside the CGA study region). Thus, the analyses here reveal no evidence for successful gene flow into the CGA population.  

Overall, results from $F_{ST}$, PCA, and `admixture` are quite similar and in agreement with previous findings (*e.g.*, from [Hooker et al. 2019](https://doi.org/10.2192/URSUS-D-18-00025.1)). Back bears in central GA are more genetically differentiated from bears in north and south GA than the north and south GA populations are from one another (despite these populations being described as *Ursus americanus americanus* and *U. a. floridanus*, respectively). This is most likely due to strong genetic drift in the small central GA population, rather than a long history of genetic distinctiveness.  


---




