---
title: "Effective Population Size"
author: "John D. Robinson"
date: "`r Sys.Date()`"
output: 
  pdf_document: 
    toc: true
linkcolor: blue
---

# Estimating Effective Population Size ($N_e$) 
In this markdown, we'll use the linkage disequilibrium (LD) method ([Hill 1981](https://doi.org/10.1017/S0016672300020553)) to estimate effective population size of the three black bear populations in Georgia using the samples processed via capture-seq. For this analysis, we'll use the same set of SNPs (`combined.vcf`) analyzed for genetic diversity (see `2_diversity`) and population differentiation (see `3_structure`). Note that we'll also avoid using intrachromosomal comparisons when calculating $r^2$ for these estimates ([Waples et al. 2016](https://doi.org/10.1038/hdy.2016.60)).  

The software `NeEstimator v2` can be downloaded and installed from [this GitHub repository](https://github.com/bunop/NeEstimator2.X).  

## Create a `GENEPOP` file (from scratch) 
The specifications of this format can be found [here](https://genepop.curtin.edu.au/help_input.html).  

First, generate a file with genotypes for all samples in the `combined.vcf` file in `012` format...  
```{r engine="bash", comment = ""}
ulimit -n 3000
vcftools --vcf combined.vcf --out combined --012
```

Then, recreate the population-specific `snps` objects we used in `19_diversity`.  
```{r makeSNPs}
#Load metadata file
samps <- read.csv("../input/sampleMetaData_cloneID.csv", header = TRUE)

#Build a popmap file to map individual samples back to their population
popmap <- read.table("combined.012.indv", header = FALSE)
names(popmap) <- "vcfID"
popmap$region <- popmap$ugaID <- popmap$year <- NA
for(i in 1:nrow(popmap)) {
  if(strsplit(popmap$vcfID[i], split="_")[[1]][1] == "UGAh") {
    popmap$region[i] <- samps$region[which(samps$vcfID == 
                                             gsub("h", "", popmap$vcfID[i]))]
    popmap$ugaID[i] <- samps$ugaID[which(samps$vcfID ==
                                           gsub("h", "", popmap$vcfID[i]))]
    popmap$year[i] <- samps$year[which(samps$vcfID == 
                                         gsub("h", "", popmap$vcfID[i]))]
  } else {
    popmap$region[i] <- samps$region[which(samps$vcfID == 
                                             popmap$vcfID[i])]
    popmap$ugaID[i] <- samps$ugaID[which(samps$vcfID == popmap$vcfID[i])]
    popmap$year[i] <- samps$year[which(samps$vcfID == popmap$vcfID[i])]
  }

  if(popmap$region[i] %in% c("CGA", "CGA-H")) {
    if(strsplit(popmap$vcfID[i], split="_")[[1]][1] == "UGAh") {
      popmap$region[i] <- "CGA12"
    } else {
      popmap$region[i] <- "CGA23"
    }
  }
}

snps <- read.table("combined.012", header = FALSE)
dim(snps) #rows are individuals, columns are SNPs
dim(popmap)
#Missing data are coded as -1's

#Create subset datasets for each "population" in our dataset
# 1. The focal population in central GA sampled in this project - `CGA23`
# 2. Historic samples from central GA (from ~2012) - `CGA12`
# 3. The north GA population - `NGA`
# 4. The south GA population - `SGA`

#Make CGA12 & CGA23 snp datasets
CGA12snps <- snps[which(popmap$region == "CGA12"),]
CGA12pop <- popmap[which(popmap$region == "CGA12"),]
CGA23snps <- snps[which(popmap$region == "CGA23"),]
CGA23pop <- popmap[which(popmap$region == "CGA23"),]
nrow(CGA12snps)
nrow(CGA23snps)

NGAsnps <- snps[popmap$region == "NGA",]
NGApop <- popmap[popmap$region == "NGA",]
nrow(NGAsnps)

SGAsnps <- snps[popmap$region == "SGA",]
SGApop <- popmap[popmap$region == "SGA",]
nrow(SGAsnps)
```

There were four bears with NGA ancestry in our CGA sample. These were offspring of a mother translocated to CGA from NW GA (see `3_structure` for more details). This artificial mixture could bias our estimates of $N_e$ downwards (via addition of mixture LD). Therefore, we'll remove these four individuals from the dataset before proceeding with the estimation.   
```{r dropBears}
#Remove NGA bears from CGA sample
dropbears <- which(CGA23pop$ugaID %in% c("74-B","73-B","83-RK","89-RK"))
length(dropbears)
CGA23pop <- CGA23pop[-dropbears,]
CGA23snps <- CGA23snps[-dropbears,]
dim(CGA23snps)
dim(CGA23pop)
```

Now create an input file in `GENEPOP` format.  
```{r makeGenepop}
#Create GENEPOP input
loclist <- paste0("SNP",1:ncol(snps[,-1]))
makeGP <- function(x) {
  x[which(x == "0")] <- "0101"
  x[which(x == "1")] <- "0102"
  x[which(x == "2")] <- "0202"
  x[which(x == "-1")] <- "0000"
  return(x)
}

#For each of the 4 populations...
#CGA 2012
CGA12gp <- apply(CGA12snps[,-1], 2, FUN=makeGP)
CGA12gp <- cbind(paste0(CGA12pop$ugaID,","), CGA12gp)

#CGA 2023
CGA23gp <- apply(CGA23snps[,-1], 2, FUN=makeGP)
CGA23gp <- cbind(paste0(CGA23pop$ugaID,","), CGA23gp)

#NGA
NGAgp <- apply(NGAsnps[,-1], 2, FUN=makeGP)
NGAgp <- cbind(paste0(NGApop$ugaID,","), NGAgp)

#SGA
SGAgp <- apply(SGAsnps[,-1], 2, FUN=makeGP)
SGAgp <- cbind(paste0(SGApop$ugaID,","), SGAgp)


cat("NeEstimator - all pops", file = "CGA_GPfile.txt")
cat("\n", file = "CGA_GPfile.txt", append = T)
cat(paste0(loclist, collapse = ","), file = "CGA_GPfile.txt", append = T)
cat("\n", file = "CGA_GPfile.txt", append = T)
cat("POP", file = "CGA_GPfile.txt", append = T)
cat("\n", file = "CGA_GPfile.txt", append = T)
write.table(CGA12gp, file = "CGA_GPfile.txt", 
            append = T, quote = F, sep = " ", row.names = F, col.names = F)
cat("POP", file = "CGA_GPfile.txt", append = T)
cat("\n", file = "CGA_GPfile.txt", append = T)
write.table(CGA23gp, file = "CGA_GPfile.txt", 
            append = T, quote = F, sep = " ", row.names = F, col.names = F)
cat("POP", file = "CGA_GPfile.txt", append = T)
cat("\n", file = "CGA_GPfile.txt", append = T)
write.table(NGAgp, file = "CGA_GPfile.txt", 
            append = T, quote = F, sep = " ", row.names = F, col.names = F)
cat("POP", file = "CGA_GPfile.txt", append = T)
cat("\n", file = "CGA_GPfile.txt", append = T)
write.table(SGAgp, file = "CGA_GPfile.txt", 
            append = T, quote = F, sep = " ", row.names = F, col.names = F)


#Now create the SNP info file with chromosome information
cpos <- read.table("combined.012.pos", header = FALSE)
snpkey <- data.frame(CHROM=cpos[,1], SNP=loclist)
write.table(snpkey, "SNPkey.txt", 
            quote = FALSE, row.names = FALSE, col.names = TRUE)
```

## Run `NeEstimator`
The analysis requires `info` and `option` files as well. These were created in a text editor.  
```{r engine="bash", comment = ""}
cat ../input/CGA_info.txt

cat ../input/CGA_option.txt
```

Now run NeEstimator with these files, this doesn't take too terribly long...  

*n.b.*, this chunk did not work on the UGA cluster. Repeated errors that NeEstimator exited with status 134. However, the program generates the expected output. Setting `eval=FALSE` in this chunk header and using output from the first submission (earlier today) in the generation of the markdown.  
```{r NeStep}
NeEst="/home/jdrobins/Software/NeEstimator2.X/Ne2L"
$NeEst i:../input/CGA_info.txt o:../input/CGA_option.txt
```

## $N_e$ estimates
Now let's look at the output...  
\tiny
```{r engine='bash', comment=''}
cat CGA_NexLD.txt
```

---
