---
title: "MHC Diversity"
author: "John D. Robinson"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    toc: true
linkcolor: blue
---

# Characterizing Diversity at MHC Loci
The major histocompatibility complex (MHC) is an important component of the immune response in vertebrates. Numerous previous studies have documented evidence of balancing selection acting on the region. Given its role in adaptive immunity, levels of genetic diversity within the region are especially important to assess in populations of conservation concern. With this in mind, the probe panel used in our project included several probes designed to target MHC loci. The following references were used during [probe design](https://docs.google.com/spreadsheets/d/10tIu7FyFX5RChB88E4G4GSgiBU7_5Ciymitw1_f5u60/edit?usp=sharing) for this project.

1. [Goda et al. (2010)](https://doi.org/10.1111/j.1399-0039.2010.01528.x) - MHC class II *DRB* genes in Ursidae, evidence for balancing selection on the locus, likely duplicated (3 alleles detected) in *U. americanus*.
2. [Yasukochi et al. (2010)](https://doi.org/10.1266/ggs.85.147) - successfully sequenced expressed portions of the MHC class II *DQB* gene in *U. thibetanus*, identified exon 2
3. [Yasukochi et al. (2012)](https://doi.org/10.1186/1471-2148-12-230) - assess diversity in exon 2 of *DQB* in *U. thibetanus japonicus*. Evidence for balancing selection on the locus, likely reduced diversity due to a bottleneck.
4. [Kuduk et al. (2012)](https://doi.org/10.1186/1471-2148-12-197) - surveyed diversity in MHC class I, *DRB*, *DQA*, *DQB* (latter 3 are MHC class II genes) in *U. arctos*. High allelic diversity at MHC class I and *DRB*. The rate of non-synonymous substitution ($d_N$) was higher than the rate of synonymous substitution ($d_S$) at antigen binding sites in *DRB* and *DQB* (evidence for selection).
5. [Weber et al. (2013)](https://doi.org/10.1111/acv.12045) - low diversity at *DQA* and *DQB* in *U. maritimus*.
6. [Ishibashi et al. (2017)](https://doi.org/10.1007/s10592-016-0897-3) - further loss of diversity in exon 2 of *DQB* in *U. thibetanus japonicus*, compared to Yasukochi et al. (2012).
7. [Jia et al. (2023)](https://doi.org/10.1007/s13364-023-00685-3) - focused on exon 2 of *DQB* in *U. thibetanus*, high allelic diversity, evidence for trans-species polymorphism. $d_N > d_S$ at codons in peptide-binding regions.

Sequence accessions from these studies were used to develop capture-seq probes that targeted MHC genes. 

## Looking for targeted MHC markers
We started by searching for the accession numbers identified above in the *Ursus americanus* reference genome (using "BLAST the reference genome" from the [NCBI taxonomy page for *U. americanus*](https://www.ncbi.nlm.nih.gov/datasets/taxonomy/9643/)). We then looked for matching regions from the reference genome in the raw VCF from RAPiD Genomics using `grep` to pull out the relevant scaffold (`CHROM` column in the VCF). Results of these searches can be reviewed [in this spreadsheet](https://docs.google.com/spreadsheets/d/1T_he9I20vcJy4ewPTj7cqslRPAGYpTr26_BtK7w2jYs/edit?usp=sharing) (see "Second Attempt" tab for the most recent set of results).  

Only two of the regions identified by BLAST were present in our raw VCF:

1. `NW_025577149.1:1531301-1531698` - identified by several studies, annotated in the reference genome as *DQB*, looking more closely with [`snpEff`](https://pcingola.github.io/SnpEff/) ([Cingolani et al. 2012](http://dx.doi.org/10.4161/fly.19695)), we can narrow exon 2 down to `NW_025577149.1:1531458-1531646`.
2. `NW_025577149.1:1535166-1535535` - close match to a portion of the cDNA sequence from Yasukochi et al. (2010), appears to be exon 6 of *DQB* according to the NCBI genome data viewer. Looking more closely with  [`snpEff`](https://pcingola.github.io/SnpEff/) ([Cingolani et al. 2012](http://dx.doi.org/10.4161/fly.19695)), most of this chunk is 3' UTR.

We'll extract the coding sequences from these windows and summarize diversity in each of the populations at SNPs within these regions (including the number of non-synonymous substitutions, as inferred by the `snpEff` annotation).  

---

# Extracting MHC markers
This chunk of code creates `.bed` files for extracting the markers around exons 2 and 6 with `vcftools`. We'll start with the `PopGen0.recode.vcf` file that served as the beginning point for our filtering in `1_filterPopGen`. This file has had duplicate SNPs, half-calls, indels, and duplicate genotypes removed, but no other filters applied. We also want to drop samples with high levels of excess heterozygosity, so we'll retain only samples present in `PopGen.vcf`. First, we extract SNPs within these regions to identify boundaries of the translated sequence, later we will apply some filters for SNP quality and genotype depth.  
```{r engine="bash", comment =""}
grep ^#CHROM ../snpfiltering/PopGen.vcf | tr '\t' '\n' | grep ^UGA > keepUs.txt
#DQB exon 2: NW_025577149.1:1531301-1531698
echo 'CHROM START END' > DQBex2.bed
echo 'NW_025577149.1 1531300 1531699' >> DQBex2.bed
vcftools --vcf ../snpfiltering/PopGen0.recode.vcf \
          --bed DQBex2.bed --keep keepUs.txt --out DQBex2 --recode

#DQB exon 6: NW_025577149.1:1535166-1535535
echo 'CHROM START END' > DQBex6.bed
echo 'NW_025577149.1 1535165 1535536' >> DQBex6.bed
vcftools --vcf ../snpfiltering/PopGen0.recode.vcf \
          --bed DQBex6.bed --keep keepUs.txt --out DQBex6 --recode
```

# Annotate SNPs in these regions
We will annotate the SNPs using `snpEff` ([Cingolani et al. 2012](http://dx.doi.org/10.4161/fly.19695)), downloaded from [this page](https://pcingola.github.io/SnpEff/) or (in the case of our virtual environment) installed via `conda` from the `bioconda` channel. In order to use this software to annotate SNPs in our vcf, we first need to configure a new genome in the `snpEff` database. We followed instructions from the [`snpEff` GitHub page](https://pcingola.github.io/SnpEff/snpeff/build_db/#step-1-configure-a-new-genome) for these steps.  

*n.b.*, the reference genome configuration was completed in an interactive session before creating this markdown (`eval=FALSE` for the chunks below).  

## `snpEff` setup
### Download reference genome 
First, we downloaded the reference genome files from [NCBI's *U. americanus* taxonomy page](https://www.ncbi.nlm.nih.gov/datasets/taxonomy/9643/) and renamed them as follows:

i. `sequences.fa` - the reference genome in FASTA format ("Genome sequences" from [NCBI download](https://www.ncbi.nlm.nih.gov/datasets/taxonomy/9643/))
ii. `genes.gtf` - annotation information in GTF format ("Annotation features (GTF)" from [NCBI download](https://www.ncbi.nlm.nih.gov/datasets/taxonomy/9643/))

These files were saved into a `data/jax_bbear` subdirectory in our virtual environment:  
`veGAbears/share/snpeff-5.2-1/data/jax_bbear`    

### Configure the `snpEff` database
We then edit the configuration file (`veGAbears/share/snpeff-5.2-1/snpEff.config`) to include the following lines (in the `Third party databases` section of the `.config` file):
```{bash eval=FALSE}
# Black bear genome, jax lab
jax_bbear.genome : Black Bear
```

After that, we create the new database using `snpEff` (*n.b.*, `eval=FALSE` for this chunk). Note that we are skipping two checks that can be performed by `snpEff` with the `-noCheckCds` and `-noCheckProtein` flags.  
```{r configGenome, eval=FALSE}
snpEff build -gtf22 -v jax_bbear -noCheckCds -noCheckProtein
```

### Test configuration
The step above takes a few minutes (~7 min, when run interactively), but prints lots of information on progress. Once it is complete, we can check that the genome was configured successfully with...  
```{r engine="bash", comment =""}
snpEff dump jax_bbear | head 
```

## Annotate SNPs from *DQB*  
Now annotate the two *DQB* exons with `snpEff`. This will help us identify the boundaries of the exon and avoid splice region variants. It will also give us information on the nature of the mutations underlying each SNP (*i.e.*, missense, nonsense, etc.).  

### Exon 2
During annotation, `snpEff` adds a line describing the information in the annotation field to the header of the `.vcf` file:
\tiny
```{r engine="bash", comment = ""}
snpEff jax_bbear DQBex2.recode.vcf > DQBex2.ann.vcf
grep ANN DQBex2.ann.vcf | head -1
```
\normalsize

Here are the annotations for exon 2:  
\tiny
```{r engine="bash", comment = ""}
grep -v ^# DQBex2.ann.vcf | cut -f 1,2,8
```
\normalsize

The translated portion of exon 2 ranges from 1531403 to 1531659.  

### Exon 6
\tiny
```{r engine="bash", comment=""}
snpEff jax_bbear DQBex6.recode.vcf > DQBex6.ann.vcf
grep -v ^# DQBex6.ann.vcf | cut -f 1,2,8
```
\normalsize

Most of the exon 6 chunk in our dataset is in a 3' untranslated region (aside from 1535177 to 1535179).  


## Extract *only* the exons
Note that we lose the annotation information when we subset to these regions, so we'll have to annotate the SNPs retained after filtering the `.vcf` again later. Our strategy will be to i) identify translated portions of the exons (done above), ii) trim to those regions, iii) apply filters for quality, genotyping depth, etc., iv) phase the filtered SNPs, and v) re-annotate the loci that pass our filters.  

### Exon 2
```{r engine="bash", comment = ""}
echo 'CHROM START END' > DQBex2Only.bed
echo 'NW_025577149.1 1531402 1531659' >> DQBex2Only.bed
vcftools --vcf DQBex2.recode.vcf --bed DQBex2Only.bed \
          --out DQBex2.sub --recode
```

### Exon 6   
```{r engine="bash", comment=""}
echo 'CHROM START END' > DQBex6Only.bed
echo 'NW_025577149.1 1535176 1535179' >> DQBex6Only.bed
vcftools --vcf DQBex6.recode.vcf --bed DQBex6Only.bed \
          --out DQBex6.sub --recode
```

## Filter the dataset
Note that *missing data are imputed* during the statistical phasing in `beagle`. With that in mind, we'll impose a fairly restrictive filter on missing data. This will minimize the need to impute genotypes.  

### Filtering *DQB* Exon 2
Now filter for genotype quality ($Q \ge 30$), genotype depth ($DP \ge 10$ reads), and minor allele count ($\mathrm{mac} \ge 2$. The last of these filters will remove singleton variants, so we will underestimate haplotype diversity in the analyses below. However, the accuracy of statistical phasing is low for singleton SNPs, so the information lost is not too great. Furthermore, we are removing singletons across the entire dataset (rather than population-by-population).  
```{r engine="bash", comment = ""}
#Exon 2
#Site Quality, Genotyping Depth, Missingness, Minor Allele Count
vcftools --vcf DQBex2.sub.recode.vcf --out DQBex2_filt1 \
          --minQ 30 --minDP 10 --mac 2 \
          --remove-filtered-all --recode

#Check frequencies
vcftools --vcf DQBex2_filt1.recode.vcf --out DQBex2_filt1 \
          --freq
```

Now ID bi-allelic sites from the frequencies file...  
```{r biallelic}
#Retain only bi-allelic loci
snpfreq <- read.table("DQBex2_filt1.frq", fill = TRUE)
snpfreq <- snpfreq[-1,]
colnames(snpfreq) <- c("CHROM","POS","N_ALLELES","N_CHR", "A1","A2","A3")
snpfreq

biallelic <- snpfreq[which(snpfreq$N_ALLELES == 2),]
nrow(biallelic)

fixed.sites <- c(grep(":1", biallelic$A1), 
                 grep(":1", biallelic$A2), 
                 grep(":1", biallelic$A3))

if(length(fixed.sites) > 0) {
  fixed.sites
  biallelic[fixed.sites,]
  confirmed.biallelic <- biallelic[-fixed.sites,c(1,2)]  
} else {
  confirmed.biallelic <- biallelic
}

nrow(confirmed.biallelic)
write.table(confirmed.biallelic, file = "DQBex2_biallelic.txt",
            quote = FALSE, row.names = FALSE, col.names = FALSE)
```

And filter to keep only those sites...  
```{r engine="bash", comment=""}
#Biallelic SNPs
vcftools --vcf DQBex2_filt1.recode.vcf --out DQBex2_filt2 \
          --positions DQBex2_biallelic.txt \
          --remove-filtered-all --recode
          
#Check missing data levels
vcftools --vcf DQBex2_filt2.recode.vcf --out DQBex2_filt2 \
          --missing-indv
```

Identify individuals to remove based on missing data...  
```{r missingEx2}
ex2miss <- read.table("DQBex2_filt2.imiss", header = TRUE)
range(ex2miss$N_MISS)
hist(ex2miss$N_MISS, breaks=10)

keepind <- ex2miss$INDV[which(ex2miss$N_MISS <= 5)]

write.table(keepind, "ex2_keepind.txt", 
            quote = FALSE, row.names = FALSE, col.names = FALSE)
```

Keep only individuals with little missing data ($\mathrm{N_{MISS}} \le 5$)...  
```{r engine="bash", comment = ""}
vcftools --vcf DQBex2_filt2.recode.vcf --out DQBex2_final \
          --keep ex2_keepind.txt --recode
```

### Filtering *DQB* Exon 6
Filter for missing data, genotype quality, genotype depth, and so on...  
```{r engine="bash", comment = ""}
#Site Quality, Genotyping Depth, Missingness, Minor Allele Count
vcftools --vcf DQBex6.sub.recode.vcf --out DQBex6_filt \
          --minQ 30 --minDP 10 --mac 2 \
          --remove-filtered-all --recode
```

This second region of *DQB* is largely composed of 3' UTR. The 2 SNPs that were not in the 3' UTR had low quality (both) or three alleles (one of the two). They do not pass our filters and we are left with only exon 2. See below for SNP qualities...  
```{r engine="bash", comment = ""}
grep -v ^## DQBex6.sub.recode.vcf | cut -f 1,2,5,6
```

## Phase *DQB* exon 2 SNPs with `beagle`
Now we use `beagle` to statistically phase the SNP loci in exon 2 of *DQB* and *impute genotypes* in samples with missing data. Since the phasing of SNPs involves imputation, we will only calculate haplotype-based measures of diversity from the phased dataset. For SNP-based diversity statistics, we'll return to the unphased dataset with missing data present.  
```{r engine="bash", comment = ""}
beagle gt=DQBex2_final.recode.vcf out=DQBex2phased
```

## Re-annotate phased VCF
Run `snpEff` again with the phased dataset...  
```{r engine="bash", comment = ""}
snpEff jax_bbear DQBex2phased.vcf.gz > DQBex2phased.ann.vcf
```

As a reminder, the annotation field has the following information...  
\tiny
```{r engine="bash", comment=""}
grep ANN DQBex2phased.ann.vcf | head -1
```
\normalsize

Here are the annotations for the exon 2 SNPs that pass our filters:  
\tiny
```{r engine="bash", comment = ""}
grep -v ^# DQBex2phased.ann.vcf | cut -f 1,2,8
```
\normalsize 

Now create a `.csv` file with annotations for the final set of SNPs...  
```{r grabAnns}
CPA <- data.frame(CHROM=system("grep -v ^# DQBex2phased.ann.vcf | cut -f 1", 
                               intern = TRUE),
                  POS=system("grep -v ^# DQBex2phased.ann.vcf | cut -f 2", 
                             intern = TRUE),
                  ANN=system("grep -v ^# DQBex2phased.ann.vcf | cut -f 8", 
                             intern = TRUE))
write.csv(CPA, file="DQBex2ann.csv", quote = FALSE, row.names = FALSE)
```

## Extract haplotype data from phased VCF
Use bcftools to extract haplotypes from the phased `.vcf` file...  
```{r engine="bash", comment = ""}
bcftools convert --hapsample --vcf-ids DQBex2phased.ann.vcf -o DQBex2
gunzip DQBex2.hap.gz
```

Now read the output into R...  
```{r readHap}
#Read the sample IDs
sampIDs <- read.table("DQBex2.samples", header = TRUE)
sampIDs <- sampIDs[-1,]
dim(sampIDs)

#Read the hap data
hap <- read.table("DQBex2.hap", header = FALSE)
dim(hap)
hap[1:10,1:10]
#Name the columns
colnames(hap)[1:5] <- c("CHROM", "ID", "POS", "REF","ALT")
al1col <- seq(from=6, by=2, length.out=nrow(sampIDs))
for(i in 1:length(al1col)) {
  colnames(hap)[al1col[i]] <- paste0(sampIDs[i,1],".1")
  colnames(hap)[(al1col[i]+1)] <- paste0(sampIDs[i,1], ".2")
}

#Now join the genotypes into a haplotype string and save it to a data.frame
samps <- read.csv("../input/sampleMetaData_cloneID.csv", header = TRUE)
hap.out <- data.frame(vcfID=sampIDs[,1], 
                      pop = NA, survivor=NA,
                      allele1 = NA, allele2 = NA)
for(i in 1:nrow(hap.out)) {
  hap.out$pop[i] <- samps$region[which(samps$vcfID == hap.out$vcfID[i])]
  hap.out$allele1[i] <- paste0(hap[,which(colnames(hap) == 
                                            paste0(hap.out$vcfID[i], ".1"))], 
                               collapse="")
  hap.out$allele2[i] <- paste0(hap[,which(colnames(hap) == 
                                            paste0(hap.out$vcfID[i],".2"))], 
                               collapse="")
  hap.out$survivor[i] <- ifelse(
    length(unique(
      samps$region[which(samps$finalClone == 
                           samps$finalClone[which(samps$vcfID == 
                                                    hap.out$vcfID[i])])]
      )) == 1, 
    FALSE, TRUE)
}
hap.out[1:10,]
```

# Haplotype-based summaries 
## Haplotype richness ($K$), haplotype diversity ($\hat{H}$), and observed heterozygosity ($H_o$)
First, working with the number of unique haplotypes $K$ and [Nei's (1987)](https://doi.org/10.7312/nei-92038) haplotype diversity, $\hat{H}$:
$$\hat{H}=\left( \frac{n}{n-1} \right) \left( 1-\sum_{i=1}^k x_i^2 \right)$$
where $n$ is the number of sequenced chromosomes, $k$ is the number of unique haplotypes detected, and $x_i$ is the frequency of the $i^{th}$ haplotype.  
```{r hapStats}
#Create population-specific objects
# For CGA23 and CGA12, don't forget to include surviving individuals!
CGA23haps <- hap.out[which(hap.out$pop == "CGA" | hap.out$survivor == TRUE),]
nrow(CGA23haps)
CGA12haps <- hap.out[which(hap.out$pop == "CGA-H" | hap.out$survivor == TRUE),]
nrow(CGA12haps)
NGAhaps <- hap.out[which(hap.out$pop == "NGA"),]
nrow(NGAhaps)
SGAhaps <- hap.out[which(hap.out$pop == "SGA"),]
nrow(SGAhaps)

#Number of unique haplotypes...
length(unique(c(hap.out$allele1, hap.out$allele2)))
length(unique(c(CGA23haps$allele1, CGA23haps$allele2)))
length(unique(c(CGA12haps$allele1, CGA12haps$allele2)))
length(unique(c(NGAhaps$allele1, NGAhaps$allele2)))
length(unique(c(SGAhaps$allele1, SGAhaps$allele2)))

#Haplotype diversity (Nei 1987)
hapdiv <- function(haps) {
  n <- 2*nrow(haps)
  haptab <- table(c(haps$allele1, haps$allele2))
  freqs <- haptab/sum(haptab)
  H <- (n/(n-1))*(1-sum(freqs^2))
  return(H)
}

#Nei's haplotype diversity
hapdiv(CGA23haps)
hapdiv(CGA12haps)
hapdiv(NGAhaps)
hapdiv(SGAhaps)

#Observed heterozygosity (haplotype-based)
sum(CGA23haps$allele1 != CGA23haps$allele2)/nrow(CGA23haps)
sum(CGA12haps$allele1 != CGA12haps$allele2)/nrow(CGA12haps)
sum(NGAhaps$allele1 != NGAhaps$allele2)/nrow(NGAhaps)
sum(SGAhaps$allele1 != SGAhaps$allele2)/nrow(SGAhaps)
```
Encouragingly, the central GA population seems to have retained quite a bit of diversity at exon 2 of *DQB*!  

## Haplotype-sharing across populations
We're also interested in how much differentiation there is in haplotype frequencies across the three black bear populations in GA.  
```{r sharedHaps}
length(unique(c(hap.out$allele1, hap.out$allele2))) #Total unique haplotypes
pophapcnts <- data.frame(hapID=c(1:length(unique(c(hap.out$allele1, 
                                                    hap.out$allele2)))),
                          DQBex2Hap=unique(c(hap.out$allele1, hap.out$allele2)),
                          NGA=NA,
                          CGA12=NA,
                          CGA23=NA,
                          SGA=NA)
for(haplo in pophapcnts$hapID) {
  #print(hap)
  #NGA frequencies
  pophapcnts$NGA[haplo] <- 
    sum(c(hap.out$allele1[hap.out$pop=="NGA"],
          hap.out$allele2[hap.out$pop=="NGA"]) == 
          pophapcnts$DQBex2Hap[haplo])
  
  #SGA frequencies
  pophapcnts$SGA[haplo] <- 
    sum(c(hap.out$allele1[hap.out$pop=="SGA"],
          hap.out$allele2[hap.out$pop=="SGA"]) == 
          pophapcnts$DQBex2Hap[haplo]) 
  
  #CGA12 frequencies (CGA-H)
  pophapcnts$CGA12[haplo] <- 
    sum(c(hap.out$allele1[hap.out$pop=="CGA-H"],
          hap.out$allele2[hap.out$pop=="CGA-H"]) == 
          pophapcnts$DQBex2Hap[haplo]) 
  
  #CGA23 frequencies (CGA)
  pophapcnts$CGA23[haplo] <- 
    sum(c(hap.out$allele1[hap.out$pop=="CGA"],
          hap.out$allele2[hap.out$pop=="CGA"]) == 
          pophapcnts$DQBex2Hap[haplo]) 
  
}

#Haplotype counts
pophapcnts

#Haplotype frequencies
pophapfreqs <- pophapcnts
for(popcol in c(3:6)) {
  pophapfreqs[,popcol] <- pophapfreqs[,popcol]/sum(pophapfreqs[,popcol])
}
pophapfreqs

#Highest frequency haplotype in each population?
which.max(pophapfreqs[,3])
which.max(pophapfreqs[,4])
which.max(pophapfreqs[,5])
which.max(pophapfreqs[,6])

pophapfreqs[1:4,]
```

There are some notable differences in the frequencies of haplotypes across the populations. For instance, the most frequent haplotype in both `NGA` and `SGA` (row #1 above) is at moderate frequency in `CGA23`.  

## Visualizing differences across populations
Here we'll take a look at the *proportion of each population* that carries each of the unique *DQB* alleles identified in our dataset. This is slightly different from a plot of haplotype frequencies by population and is inspired by Figure 3a of [Grogan et al. (2017)](https://doi.org/10.1002/ece3.3317). This should help to illustrate patterns of diversity within populations and differentiation among populations at *DQB*.  
```{r barplotAF, fig.height = 9, fig.width = 4.5}
#Make an object that we can use to store the data more efficiently
# nrow is the number of individuals sampled total
# ncol is the number of unique haplotypes
DQBmat <- matrix(data = 0, nrow = nrow(hap.out), ncol = nrow(pophapcnts))
colnames(DQBmat) <- pophapcnts$hapID
for(i in 1:nrow(hap.out)) {
  hap.cols <- c(which(pophapcnts$DQBex2Hap == hap.out$allele1[i]),
                which(pophapcnts$DQBex2Hap == hap.out$allele2[i]))
  DQBmat[i,hap.cols[1]] <- DQBmat[i,hap.cols[1]]+1
  DQBmat[i,hap.cols[2]] <- DQBmat[i,hap.cols[2]]+1
  rm(hap.cols)
}
rowSums(DQBmat) #should all be 2 - 2 alleles per individual
colSums(DQBmat) #should be equal to
rowSums(pophapcnts[,-c(1,2)])
sum(colSums(DQBmat) == rowSums(pophapcnts[,-c(1,2)])) #for all 44 unique haps

#Good!

#Now make an object with the proportion of the population that carries each hap
DQBprops <- data.frame(hapID = pophapcnts$hapID, NGA=0, CGA=0, CGAH=0, SGA=0)
for(i in 1:nrow(DQBprops)) {
  DQBprops$NGA[i] <- sum(DQBmat[hap.out$pop=="NGA",i]>0)/sum(hap.out$pop=="NGA")
  DQBprops$CGA[i] <- sum(DQBmat[hap.out$pop=="CGA",i]>0)/sum(hap.out$pop=="CGA")
  DQBprops$CGAH[i] <- sum(DQBmat[hap.out$pop=="CGA-H",i]>0)/sum(hap.out$pop=="CGA-H")
  DQBprops$SGA[i] <- sum(DQBmat[hap.out$pop=="SGA",i]>0)/sum(hap.out$pop=="SGA")
}
DQBprops <- t(DQBprops[,-1])

DQBprops

barplot(DQBprops[c(4,3,2,1),ncol(DQBprops):1], beside=TRUE, horiz=TRUE, 
        names = c(ncol(DQBprops):1), las=1, cex.names=0.5, 
        xlab = "Proportion of Population", ylab = "DQB Allele", 
        col = c("lightblue","lightgreen","darkgreen","darkgoldenrod1"))
legend("bottomright", bty="n", 
       fill=c("darkgoldenrod1","lightgreen","darkgreen","lightblue"), 
       legend = c("NGA","CGA-H","CGA","SGA"))
```

## Influences of sample size
The imbalanced sample sizes across populations makes it difficult to compare diversity levels in plots like the one above. [Grogan et al. (2013)](https://doi.org/10.1002/ece3.3317) used rarefaction approaches (similar to species accumuation curves) to compare the accumulation of unique alleles with additional samples in captive and wild populations of ring-tailed lemurs (*Lemur catta*). We'll use the `specaccum` function from the R package `vegan` (with `method = "exact"`) to do something similar here. See the ["Vegan: ecological diversity" vignette](https://cran.r-project.org/web/packages/vegan/vignettes/diversity-vegan.pdf) (section 5.1, in particular) for methods that underlie `specaccum`. Kindt's exact method is the recommended approach for species accumulation curves, thus we use that approach below.  

The `specaccum` function takes a community dataset where rows are site visits (in our case, sampled individuals) and columns are species (alleles, in our case). So we can use the object created above (`DQBmat`) for this analysis. Note that we are leaving out the 2012 samples from the central GA population. The numbers there are less than half the sample size in the next most poorly sampled population, including that sample would unnecessarily constrain the accumulation curves for the comparisons of interest.  
```{r accumcurve}
library(vegan)
library(scales)

NGAmat <- DQBmat[which(hap.out$pop == "NGA"),]
CGAmat <- DQBmat[which(hap.out$pop == "CGA"),]
SGAmat <- DQBmat[which(hap.out$pop == "SGA"),]
CGAHmat <- DQBmat[which(hap.out$pop == "CGA-H"),]

NGAsac <- specaccum(NGAmat, method = "exact")
SGAsac <- specaccum(SGAmat, method = "exact")
CGAsac <- specaccum(CGAmat, method = "exact")
CGAHsac <- specaccum(CGAHmat, method = "exact")

#Truncate the plot to avoid plotting very large n in CGA 2023
plot(CGAsac, xlim = c(0,35), ci.type="polygon", lwd = 1.5,
     col = "darkgreen", ci.col = alpha("darkgreen", 0.5), ci.lty=0,
     xlab = "Number of Individuals Sampled", ylab = "Unique DQB Alleles")
plot(SGAsac, add = TRUE, ci.type = "polygon", lwd = 1.5, 
     col = "lightblue", ci.col = alpha("lightblue", 0.75), ci.lty=0)
plot(NGAsac, add = TRUE, ci.type = "polygon", lwd = 1.5, 
     col = "darkgoldenrod1", ci.col=alpha("darkgoldenrod1", 0.5), ci.lty=0)
plot(CGAHsac, add = TRUE, ci.type = "polygon", lwd = 1.5, 
     col = "lightgreen", ci.col=alpha("lightgreen",0.5), ci.lty=0)
legend("topleft", lty = 1, bty = "n", cex = 0.75,
       col = c("darkgoldenrod1","darkgreen","lightgreen","lightblue"), 
       legend = c("NGA","CGA","CGA-H","SGA"))

#Same plot, without truncation
plot(CGAsac, ci.type="polygon", lwd = 1.5,
     col = "darkgreen", ci.col = alpha("darkgreen", 0.5), ci.lty=0,
     xlab = "Number of Individuals Sampled", ylab = "Unique DQB Alleles")
plot(SGAsac, add = TRUE, ci.type = "polygon", lwd = 1.5, 
     col = "lightblue", ci.col = alpha("lightblue", 0.75), ci.lty=0)
plot(NGAsac, add = TRUE, ci.type = "polygon", lwd = 1.5, 
     col = "darkgoldenrod1", ci.col=alpha("darkgoldenrod1", 0.5), ci.lty=0)
plot(CGAHsac, add = TRUE, ci.type = "polygon", lwd = 1.5, 
     col = "lightgreen", ci.col=alpha("lightgreen",0.5), ci.lty=0)
legend("topleft", lty = 1, bty = "n", cex = 0.75,
       col = c("darkgoldenrod1","lightgreen","darkgreen","lightblue"), 
       legend = c("NGA","CGA-H","CGA","SGA"))
```

## Neighbor-Joining Tree of *DQB* 
Now we use the package `ape` to make a neighbor-joining phylogeny of *DQB*. Note that, since this tree is based on SNP data only, the branch lengths are a bit exaggerated (*i.e.*, invariant sites are not included).  
```{r DQBphylo}
#First, convert to a fake DNA sequence - gsub 0/1 to REF/ALT
seqhap <- hap[,-c(1:5)]
for(snp in 1:nrow(hap)) {
  seqhap[snp,] <- gsub("0", hap$REF[snp], seqhap[snp,])
  seqhap[snp,] <- gsub("1", hap$ALT[snp], seqhap[snp,])
}

library(ape)
DQBaln <- as.alignment(seqhap)
DQBbin <- as.DNAbin(DQBaln)
dmat <- dist.dna(DQBbin, model = "N")
NJtree <- nj(dmat)
#Relabel tips as pops
poplabels <- tipcols <- c()
for(i in 1:length(NJtree$tip.label)) {
  poplabels[i] <- hap.out$pop[which(hap.out$vcfID == 
                                      strsplit(NJtree$tip.label[i], 
                                               split=".", 
                                               fixed=TRUE)[[1]][1])]
}
NJtree$tip.label <- poplabels
  
tipcols[which(poplabels == "NGA")] <- "darkgoldenrod1"
tipcols[which(poplabels == "SGA")] <- "lightblue"
tipcols[which(poplabels == "CGA")] <- "darkgreen"
tipcols[which(poplabels == "CGA-H")] <- "lightgreen"

#Plot the tree
par(mar=c(0,0,0,0))
plot(NJtree, tip.col=tipcols, cex=0.5)
legend("topright", bty = "n", cex = 0.75,
       legend = c("NGA","CGA-H","CGA","SGA"), 
       text.col=c("darkgoldenrod1","lightgreen","darkgreen","lightblue"))
```

## Haplotype network of *DQB* 
Now a haplotype network using tools from the `pegas` package. The below code closely follows the vignette from the `pegas` package [Plotting Haplotype Networks with `pegas`](https://cran.r-project.org/web/packages/pegas/vignettes/PlotHaploNet.pdf).  
```{r DQBhapNet}
library(pegas)
library(network)

#Extract unique haplotypes
uniq.haps <- haplotype(DQBbin)
uniq.haps
#Distances among haplotypes - # of bp differences
hap.dist <- dist.dna(uniq.haps, model="N")
#Randomized minimum spanning tree for the haplotype network
hapnet <- rmst(hap.dist)
hapnet
#Record the counts of each haplotype for the plot
sz <- summary(uniq.haps)
sz
hapnet.labels <- attr(hapnet,"labels")
hapnet.labels
sz <- sz[hapnet.labels]
sz
pie.size <- sqrt(sz/pi) 
#Scale by sqrt(count / pi) so AREA of circles are proportional to numbers

#Compute frequencies of haplotypes for each population
pop.freq <- haploFreq(DQBbin, fac=poplabels, haplo=uniq.haps)
pop.freq <- pop.freq[hapnet.labels,]
pop.freq

#Use the network package to establish a reasonable layout for the network
# this solution is from the help for ?plot.haploNet & this post:
# https://stat.ethz.ch/pipermail/r-sig-genetics/2022-July/000237.html
netxy <- plot(as.network(hapnet))
xy <- list(x=netxy[,1], y=netxy[,2])

par(mar=c(0,0,0,0))
plot(hapnet, size=pie.size, xy=xy, pie=pop.freq, 
     labels=FALSE, show.mutation=1, lwd=0.5, 
     bg=c("darkgreen","lightgreen","darkgoldenrod1","lightblue"))
legend("topleft", bty="n", cex = 0.75,
       fill=c("darkgoldenrod1","lightgreen","darkgreen","lightblue"),
       legend=c("NGA","CGA-H","CGA","SGA"))
```



```{r hapStatsOUT, echo = FALSE}
out <- data.frame(NGA = nrow(NGAhaps),
                  CGA12 = nrow(CGA12haps),
                  CGA23 = nrow(CGA23haps),
                  SGA = nrow(SGAhaps))
temp <- data.frame(NGA = length(unique(c(NGAhaps$allele1, NGAhaps$allele2))),
                  CGA12 = length(unique(c(CGA12haps$allele1, CGA12haps$allele2))),
                  CGA23 = length(unique(c(CGA23haps$allele1, CGA23haps$allele2))),
                  SGA = length(unique(c(SGAhaps$allele1, SGAhaps$allele2))))
temp2 <- data.frame(NGA = hapdiv(NGAhaps),
                  CGA12 = hapdiv(CGA12haps),
                  CGA23 = hapdiv(CGA23haps),
                  SGA = hapdiv(SGAhaps))
temp3 <- data.frame(NGA=sum(NGAhaps$allele1 != NGAhaps$allele2)/nrow(NGAhaps),
                    CGA12=sum(CGA12haps$allele1 != CGA12haps$allele2)/nrow(CGA12haps),
                    CGA23=sum(CGA23haps$allele1 != CGA23haps$allele2)/nrow(CGA23haps),
                    SGA=sum(SGAhaps$allele1 != SGAhaps$allele2)/nrow(SGAhaps))
out <- rbind(out, temp, temp2, temp3)
rownames(out) <- c("n","K_hap","H_hap","Ho_hap")
rm(temp, temp2, temp3)
```



# Predicted effects of mutations
Given the filters applied above and the lack of sequence data for exon 2 of the *DQB* gene, we will not attempt to calculate a $\frac{d_N}{d_S}$ ratio from the `snpEff` annotation. Methods to calculate these ratios typically require aligned sequences, rather than a `.vcf` file with only SNP loci. While the `snpEff` annotation does tell us about the number of non-synonymous substitutions, it does not tell us about the possible number of mutations at each site that would have led to synonymous and non-synonymous changes. Thus, we will focus on a simple summary of the numbers of synonymous and non-synonymous mutations that underlie the polymorphic SNP loci in each population.  

First, create `.vcf` files for each population.  
```{r makeMHCkeep}
write.table(CGA12haps$vcfID, file = "CGA12mhckeep.txt", 
            quote = FALSE, 
            row.names = FALSE,
            col.names = FALSE)
write.table(CGA23haps$vcfID, file = "CGA23mhckeep.txt", 
            quote = FALSE, 
            row.names = FALSE,
            col.names = FALSE)
write.table(NGAhaps$vcfID, file = "NGAmhckeep.txt", 
            quote = FALSE, 
            row.names = FALSE,
            col.names = FALSE)
write.table(SGAhaps$vcfID, file = "SGAmhckeep.txt", 
            quote = FALSE, 
            row.names = FALSE,
            col.names = FALSE)
```

Now extract each population into its own `.vcf` file and rename the historic samples so survivors are present in both 2012 and 2023 populations...  
```{r engine="bash", comment=""}
vcftools --vcf DQBex2_final.recode.vcf --keep CGA12mhckeep.txt \
          --recode --out DQBex2_CGA12
vcftools --vcf DQBex2_final.recode.vcf --keep CGA23mhckeep.txt \
          --recode --out DQBex2_CGA23
vcftools --vcf DQBex2_final.recode.vcf --keep NGAmhckeep.txt \
          --recode --out DQBex2_NGA
vcftools --vcf DQBex2_final.recode.vcf --keep SGAmhckeep.txt \
          --recode --out DQBex2_SGA
```

Now calculate allele frequencies for each population...  
```{r engine="bash",comment=""}
vcftools --vcf DQBex2_CGA12.recode.vcf --mac 1 --freq --out DQBex2_CGA12
vcftools --vcf DQBex2_CGA23.recode.vcf --mac 1 --freq --out DQBex2_CGA23
vcftools --vcf DQBex2_NGA.recode.vcf --mac 1 --freq --out DQBex2_NGA
vcftools --vcf DQBex2_SGA.recode.vcf --mac 1 --freq --out DQBex2_SGA
```

Read in the frequency info and annotation info...  
```{r countNS}
#Read the annotation information and files with info on variable positions
anns <- read.csv("DQBex2ann.csv", header=TRUE)
CGA12frq <- read.table("DQBex2_CGA12.frq", header=FALSE, skip=1)
colnames(CGA12frq) <- c("CHROM","POS","N_ALLELES","A1","A2")
CGA23frq <- read.table("DQBex2_CGA23.frq", header=FALSE, skip=1)
colnames(CGA23frq) <- c("CHROM","POS","N_ALLELES","A1","A2")
NGAfrq <- read.table("DQBex2_NGA.frq", header=FALSE, skip=1)
colnames(NGAfrq) <- c("CHROM","POS","N_ALLELES","A1","A2")
SGAfrq <- read.table("DQBex2_SGA.frq", header=FALSE, skip=1)
colnames(SGAfrq) <- c("CHROM","POS","N_ALLELES","A1","A2")

#Extract polymorphic positions for each pop
CGA12anns <- anns[which(anns$POS %in% CGA12frq$POS),3]
CGA23anns <- anns[which(anns$POS %in% CGA23frq$POS),3]
NGAanns <- anns[which(anns$POS %in% NGAfrq$POS),3]
SGAanns <- anns[which(anns$POS %in% SGAfrq$POS),3]

#Table with mutation types
table(sapply(CGA12anns, FUN=function(x) strsplit(x, "|", fixed=TRUE)[[1]][2]))
table(sapply(CGA23anns, FUN=function(x) strsplit(x, "|", fixed=TRUE)[[1]][2]))
table(sapply(NGAanns, FUN=function(x) strsplit(x, "|", fixed=TRUE)[[1]][2]))
table(sapply(SGAanns, FUN=function(x) strsplit(x, "|", fixed=TRUE)[[1]][2]))
```

```{r compMissense, echo=FALSE}
temp <- data.frame(NGA=length(grep("missense",NGAanns)),
                   CGA12=length(grep("missense",CGA12anns)),
                   CGA23=length(grep("missense",CGA23anns)),
                   SGA=length(grep("missense",SGAanns)))
out <- rbind(out, temp)
rownames(out) <- c("n","K_hap","H_hap","Ho_hap","missense")
rm(temp)
```

---

# Summary
Diversity at SNPs in exon 2 of the MHC class II gene *DQB* is lower in central GA than in north GA (particularly when considering $H_e$, $H_o$, and allelic richness). However, haplotype diversity and observed heterozygosity are both fairly high in central GA. Overall, central GA has retained a surprising amount of diversity in exon 2 of *DQB*. Furthermore, the haplotype network and neighbor-joining tree show that variants segregating in the central GA population are not particularly tightly clustered. However, many of the variants in central GA are present at low frequencies.  
```{r kable}
knitr::kable(t(out), digits = c(0,0,3,3,0))
```

---


