---
title: "Filtering VCF for Population Genetics Analyses"
author: "John D. Robinson"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    toc: true
linkcolor: blue
---

# Filter the VCF
For the population genetics analyses below, we use a different set of filters than were used for for the individual identifications that underlie capture-recapture data in [Robinson et al. (*in review*)](https://github.com/slugsrfun/CGA_SCR). These filters allow us to retain more SNPs and maintain representation across the bins of the site frequency spectrum.  

More specifically, the filters we'll apply here include the following: 

0. Cleaning the raw VCF - remove duplicated lines, half-calls, indels
1. Quality filter at the locus level - drop the SNP locus if quality < 30
2. Maximum depth filter at the locus level - drop SNP loci if total sequencing depth is > $2\sigma$ above mean total depth
3. Min depth filter at the genotype level - don't call genotypes with < 10 reads
4. Polymorphism filter - only retain loci that have exactly two alleles
5. Remove sites with minor allele count $< 3$ and sites with $> 50$% missing data
6. Remove SNPs that are very tightly linked ($r^2 > 0.8$ within 1kb windows)
7. Remove samples with high levels of excess heterozygosity ($F_{IS} < -0.5$)


*n.b.*, steps 0 through 4 are shared with the filters applied to the dataset used to identify recaptures (see filters in the [CGA_SCR repo]((https://github.com/slugsrfun/CGA_SCR)).  

## Step 0: Cleaning the Raw VCF
Previous analyses identified a large number of recaptures in our dataset, we need to subsample the data to one sample per unique genotype. We want to retain as many loci as possible, so we'll pick the sample with the least missing data (across the full raw VCF) for each unique genotype. We also only retain samples that were processed by LGC / Rapid Genomics for population genetic analyses, given the far larger number of SNP loci available for these samples.   

### Number of Unique Genotypes
First load the data, check number of "clones" per region...  
```{r loadData}
samps <- read.csv("../input/sampleMetaData_cloneID.csv", 
                  header = TRUE)
dim(samps)

#How many were we able to assign to a clone?
sum(!is.na(samps$finalClone))
#How many samples per clone?
table(samps$finalClone)

#How many per region
table(samps$region[which(samps$sampSOURCE != "Foster" &
                           samps$GTservice == "LGC")])

#North GA
table(samps$finalClone[samps$region == "NGA" & 
                         samps$sampSOURCE != "Foster" &
                         samps$GTservice == "LGC"])
#17 has two copies in the dataset (technical replicate)
samps[which(samps$finalClone == 17),]

#South GA
table(samps$finalClone[samps$region == "SGA" &
                         samps$sampSOURCE != "Foster" &
                         samps$GTservice == "LGC"])
#83, 102 have two copies in the dataset (technical replicates)
samps[which(samps$finalClone %in% c(83,102)),]

#Central GA - historic
table(samps$finalClone[which(samps$region == "CGA-H" &
                               samps$sampSOURCE != "Foster" &
                               samps$GTservice == "LGC")])
samps[which(samps$finalClone == 44),]
#A recapture in the hair samples from 2012!

#Central GA - recent
table(samps$finalClone[which(samps$region == "CGA" &
                               samps$sampSOURCE != "Foster" &
                               samps$GTservice == "LGC")])
#Many of these have multiple copies!



#No clone should be found in multiple regions...
numregions <- c()
for(i in 1:max(samps$finalClone, na.rm = TRUE)) {
  numregions[i] <- length(unique(samps$region[which(samps$finalClone == i)]))
}
numregions
max(numregions)
which(numregions == 2)
samps$region[which(samps$finalClone == 3)]
# clone 3 has representatives in recent and historic samples!
samps$region[which(samps$finalClone == 45)]
# so does cone 45!
samps$region[which(samps$finalClone == 295)]
# so does clone 295!
```

### Indels, Half-calls, Duplicates
We need to extract some information about missing data levels to identify the best-genotyped sample from each clone. We'll use `vcftools` to do that here. We also remove half calls (`./1` and `1/.`), indels, and duplicated rows of the VCF. Note that the steps below are identical to the first phase of the `filterSCR` script...    
```{r engine="bash", comment=""}
cd /scratch/jdrobins/CGA_PopGen/snpfiltering

cp ../../LGC_BlackBears/UGA_173101/Delivery-20250620_merged/VCFs/UGA_173101_SNPs_Raw.vcf ./

#Use bcftools norm to remove duplicate SNPs
bcftools norm -d snps UGA_173101_SNPs_Raw.vcf -o filt0uniq.vcf

#Use sed to replace the half-calls (./0 and ./1) with missing data (./.)
grep ^# filt0uniq.vcf > filt0.vcf
grep -v ^# filt0uniq.vcf | sed -e 's/\.\/[0,1]/\.\/\./g' | sed -e 's/[0,1]\/\./\.\/\./g' >> filt0.vcf

#Use vcftools to remove indels
vcftools --vcf filt0.vcf --remove-indels --recode --out filt0

#create file with individual missing data
vcftools --vcf filt0.recode.vcf --out raw --missing-indv

#How many SNPs do we start with?
grep -v ^# filt0.recode.vcf | wc -l
```

### Choose Representative Samples
For each of the populations (regions), go through the missing data table and identify the `vcfID` (aka `INDV`) with the least missing data from each clone. We don't want to include fostered cubs in this dataset, drop them up here. Note that samples that were filtered out of our recapture analysis are not eligible for inclusion in the population genetic analyses - we want to be sure that we aren't including duplicated samples, so no filtered samples are allowed!  
```{r buildList}
#Remove filtered samples from consideration
dim(samps)
samps <- samps[which(samps$FILT == FALSE),]
dim(samps)


#Make a list of vcfIDs that we'll pass to vcftools to subset to keep!
keeplist <- c()

#Read in the missing data summaries generated above
miss <- read.table("raw.imiss", header = TRUE)
dim(miss)

#North GA
clonesNGA <- table(samps$finalClone[which(samps$region == "NGA" & 
                                            samps$sampSOURCE != "Foster")])
clonesNGA
length(clonesNGA)

nodat <- 0
#Pull best genotyped vcfID
for(i in 1:length(clonesNGA)) {
  tmp.IDs <- samps$vcfID[which(samps$finalClone == names(clonesNGA)[i] &
                                 samps$GTservice == "LGC")]
  if(length(tmp.IDs) > 1) {
    tmp <- miss[which(miss$INDV %in% tmp.IDs),]
    keeplist <- c(keeplist,tmp$INDV[which(tmp$F_MISS == min(tmp$F_MISS))][1])
    rm(tmp)
  } else if(length(tmp.IDs) == 1) {
    keeplist <- c(keeplist, tmp.IDs)
  } else {
    nodat <- nodat + 1
    print(paste("Clone", names(clonesNGA)[i], "not sequenced by LGC:",
                nodat,"total clones without data"))
  }
  rm(tmp.IDs)
}
length(keeplist)

#South GA
clonesSGA <- table(samps$finalClone[which(samps$region == "SGA" & 
                                            samps$sampSOURCE != "Foster")])
clonesSGA
length(clonesSGA)

for(i in 1:length(clonesSGA)) {
  tmp.IDs <- samps$vcfID[which(samps$finalClone == names(clonesSGA)[i] &
                                 samps$GTservice == "LGC")]
  if(length(tmp.IDs) > 1) {
    tmp <- miss[miss$INDV %in% tmp.IDs,]
    keeplist <- c(keeplist,tmp$INDV[which(tmp$F_MISS == min(tmp$F_MISS))][1])
    rm(tmp)
  } else if(length(tmp.IDs) == 1){
    keeplist <- c(keeplist, tmp.IDs)
  } else {
    nodat <- nodat + 1
    print(paste("Clone", names(clonesSGA)[i], "not sequenced by LGC:",
                nodat,"total clones without data"))
  }
  rm(tmp.IDs)
}
length(keeplist)

#Central GA - historic
clonesCGAH <- table(samps$finalClone[which(samps$region == "CGA-H" & 
                                            samps$sampSOURCE != "Foster")])
clonesCGAH
length(clonesCGAH)

for(i in 1:length(clonesCGAH)) {
    tmp.IDs <- samps$vcfID[which(samps$finalClone == names(clonesCGAH)[i] &
                                   samps$GTservice == "LGC")]
  if(length(tmp.IDs) > 1) {
    tmp <- miss[miss$INDV %in% tmp.IDs,]
    keeplist <- c(keeplist,tmp$INDV[which(tmp$F_MISS == min(tmp$F_MISS))][1])
    rm(tmp)
  } else if(length(tmp.IDs) == 1) {
    keeplist <- c(keeplist, tmp.IDs)
  } else {
    nodat <- nodat + 1
    print(paste("Clone", names(clonesCGAH)[i], "not sequenced by LGC:",
                nodat,"total clones without data"))
  }
  rm(tmp.IDs)
}
length(keeplist)

#Central GA - recent
clonesCGA <- table(samps$finalClone[which(samps$region == "CGA" & 
                                            samps$sampSOURCE != "Foster")])
clonesCGA
length(clonesCGA)

for(i in 1:length(clonesCGA)) {
    tmp.IDs <- samps$vcfID[which(samps$finalClone == names(clonesCGA)[i] &
                                   samps$GTservice == "LGC")]
  if(length(tmp.IDs) > 1) {
    tmp <- miss[miss$INDV %in% tmp.IDs,]
    keeplist <- c(keeplist,tmp$INDV[which(tmp$F_MISS == min(tmp$F_MISS))][1])
    rm(tmp)
  } else if(length(tmp.IDs) == 1) {
    keeplist <- c(keeplist, tmp.IDs)
  } else {
    nodat <- nodat + 1
    print(paste("Clone", names(clonesCGA)[i], "not sequenced by LGC:",
                nodat,"total clones without data"))
  }
  rm(tmp.IDs)
}

#These two numbers won't match
# some clones were not sequenced by LGC...
length(keeplist)
length(clonesCGA)+length(clonesCGAH)+length(clonesNGA)+length(clonesSGA)
#These numbers should match...
length(clonesCGA)+length(clonesCGAH)+length(clonesNGA)+length(clonesSGA) - 
  length(keeplist)
nodat

#Note that the three clones with samples in both 2012 and 2023-2024 are 
# represented twice in the list, by the same sample
length(keeplist)
length(unique(keeplist))
sum(names(clonesCGAH) %in% names(clonesCGA))
names(clonesCGAH)[which(names(clonesCGAH) %in% names(clonesCGA))]
# Clone IDs 3, 45, and 295

#Table of samples by clone
keepsamps <- c()
for(i in 1:length(keeplist)) {
  keepsamps[i] <- which(samps$vcfID == keeplist[i])
}
table(samps$finalClone[keepsamps])
max(table(samps$finalClone[keepsamps]))

#Table of samples by region - off by a few
# +2 for CGA-H & +1 for CGA
table(samps$region[keepsamps])

#Table of samples by source
table(samps$sampSOURCE[which(samps$vcfID %in% keeplist)])

#Table of samples by year
table(samps$year[which(samps$vcfID %in% keeplist)])

#Write keeplist to a text file
write.table(keeplist, file = "PG_keepers.txt", 
            quote = FALSE, 
            row.names = FALSE,
            col.names = FALSE)
```

### Extract Genotypes for Selected Samples
Most of the cleaning was done above, but here we remove duplicated samples and retain one sample per unique genotype in the dataset. After this step, the VCF file will only include samples in the `PG_keepers.txt` file generated above.  
```{r engine="bash", comment =""}
#Keep only individuals identified above
vcftools --vcf filt0.recode.vcf --out PopGen0 \
      --keep PG_keepers.txt --remove-filtered-all --recode
```

```{r echo=FALSE}
snpct <- as.numeric(system("grep -v ^# PopGen0.recode.vcf | wc -l", intern=TRUE))
```

## Step 1: Quality Filter
The first true filter we apply is a genotype quality filter (`--minQ`).  

Prior to this filter, we have `r snpct` loci in the VCF...  
```{r engine="bash", comment = ""}
#Filter sites 
# min site quality of 30
# create a new VCF
vcftools --vcf PopGen0.recode.vcf --out PopGen1 \
          --minQ 30 --remove-filtered-all --recode
```

```{r echo=FALSE}
snpct <- as.numeric(system("grep -v ^# PopGen1.recode.vcf | wc -l", intern=TRUE))
```

## Step 2: Maximum Depth Filter
Next, we remove sites that are sequenced far more deeply than other markers. These may be loci where paralagous regions of the genome have been combined. We would expect markers in such regions to show elevated diversity and deviate from Hardy-Weinberg equilibrium. There are other ways to remove these (see [McKinney et al. 2017](https://doi.org/10.1111/1755-0998.12613)), but we'll use a simple maximum (total) depth filter.

Prior to these filters, we have `r snpct` SNPs in the dataset...  
```{r engine="bash", comment=""}
#Get summaries of mean depth per site (for max depth filter) 
vcftools --vcf PopGen1.recode.vcf --site-depth --out PopGen1
```

Now in R, we can use this `filt1.ldepth` file to calculate a mean total depth (summed across individuals)...  
```{r calcMeanDepth}
ldep <- read.table("PopGen1.ldepth", header = TRUE)

mean(ldep$SUM_DEPTH)
sd(ldep$SUM_DEPTH)

#This will be our threshold... 
# nothing more than 2 sd above the mean...
cap <- mean(ldep$SUM_DEPTH) + (2*sd(ldep$SUM_DEPTH))
cap

keeper.sites <- ldep[which(ldep$SUM_DEPTH <= cap),c(1,2)]

write.table(keeper.sites, file="PG_keepersites.txt", 
            quote=FALSE, row.names = FALSE, col.names = FALSE)

```

Now apply this filter...  
```{r engine="bash", comment=""}
#Then remove loci based on this threshold...
vcftools --vcf PopGen1.recode.vcf --out PopGen2 \
          --positions PG_keepersites.txt \
          --remove-filtered-all --recode
```

```{r echo=FALSE}
snpct <- as.numeric(system("grep -v ^# PopGen2.recode.vcf | wc -l", intern=TRUE))
```

## Step 3: Min Depth for Genotype Calls
In the next filter, we remove genotype calls that are based on fewer than 10 reads. We set this threshold to maximize our confidence in individual calls. 

Prior to these filters, we have `r snpct` loci in the VCF...  
```{r engine="bash", comment=""}
#Min depth for genotype calls is 10 reads
# All loci have exactly 2 alleles
# Max missing is 90% of SNPs
vcftools --vcf PopGen2.recode.vcf --out PopGen3 \
          --minDP 10 --remove-filtered-all --recode
```

```{r echo=FALSE}
snpct <- as.numeric(system("grep -v ^# PopGen3.recode.vcf | wc -l", intern=TRUE))
```

## Step 4: Polymorphism Filter
Now we want to apply a filter to remove sites that have more or less than 2 alleles. The built in `--min-alleles` and `--max-alleles` arguments in `vcftools` do not work the way we would want them to. They ignore variation present in the VCF at that stage, and instead base the number of alleles at a marker on the original calls (*i.e.*, the number of alleles implied by the `REF` and `ALT` columns of the file). This unnecessarily removes "multiallelic" sites that become biallelic after depth filters, which amounted to several thousand SNP loci in our dataset. Instead, we'll use the `--positions` argument to specify a keeper list of SNP loci where there are two alleles.  

First, generate an allele frequency summary for the vcf using `vcftools`...  
```{r engine="bash", comment = ""}
vcftools --vcf PopGen3.recode.vcf --out PopGen3 \
          --freq
```

Now read this file into R and identify sites that have exactly two alleles in the dataset...  
```{r biallelic}
snpfreq <- read.table("PopGen3.frq", skip = 1, fill=TRUE)
colnames(snpfreq) <- c("CHROM","POS","N_ALLELES","N_CHR", "A1","A2","A3")
biallelic <- snpfreq[which(snpfreq$N_ALLELES == 2),]
nrow(biallelic)

fixed.sites <- c(grep(":1", biallelic$A1), 
                 grep(":1", biallelic$A2), 
                 grep(":1", biallelic$A3))
fixed.sites
biallelic[fixed.sites,]

confirmed.biallelic <- biallelic[-fixed.sites,c(1,2)]
nrow(confirmed.biallelic)

write.table(confirmed.biallelic, file = "biallelic.txt",
            quote = FALSE, row.names = FALSE, col.names = FALSE)
```

Now retain only SNP loci in this list ...  
```{r engine="bash", comment = ""}
vcftools --vcf PopGen3.recode.vcf --out PopGen4 \
        --positions biallelic.txt \
        --remove-filtered-all --recode
```


```{r echo=FALSE}
snpct <- as.numeric(system("grep -v ^# PopGen4.recode.vcf | wc -l", intern=TRUE))
```

## Step 5: Minor Allele Count & Missing Data
SNPs that are exceedingly rare are more likely to be false alleles called due to sequencing error. A (low) minor allele frequency filter (or a minor allele count filter) can help avoid including these markers in our dataset. Sites that are missing a majority of genotypes will also be less useful for population genetic analyses.  

Prior to these filters, we have `r snpct` loci in the VCF...  
```{r engine="bash", comment = ""}
#Remove sites with minor allele count < 3 and sites with >50% missing data
vcftools --vcf PopGen4.recode.vcf --out PopGen5 \
          --mac 3 --max-missing 0.5 \
          --remove-filtered-all --recode
```

```{r echo=FALSE}
snpct <- as.numeric(system("grep -v ^# PopGen5.recode.vcf | wc -l", intern=TRUE))
```

## Step 6: Avoiding Tight Linkage
We also filter to avoid very tightly linked markers. This filter will be skipped for analysis with `LinkNe` and `GONE` - which use linkage disequilibrium at linked markers to infer changes in population size over the recent(ish) past.  

Prior to these filters, we have `r snpct` loci in the VCF...  
```{r engine="bash", comment=''}
#Prune based on LD cutoff of r2 < 0.8
plink2 --vcf PopGen5.recode.vcf --out prune \
        --indep-pairwise 1 kb 1 0.8 \
        --set-all-var-ids @:# --silent

plink2 --vcf PopGen5.recode.vcf --out PopGen6 \
        --extract prune.prune.in --allow-extra-chr \
        --recode vcf-4.2 --set-all-var-ids @:# --silent

```

```{r echo=FALSE}
snpct <- as.numeric(system("grep -v ^# PopGen6.vcf | wc -l", intern=TRUE))
```


## Step 7: Samples with Excess Heterozygosity
Samples with very low $F$ (*i.e.*, a substantial excess of observed heterozygosity, relative to expected) may be contaminated. Here, we check for samples with $F < -0.5$ and remove them from the vcf. We then double check that remaining SNPs are still biallelic before generating our final output.  

First, we need to identify the samples associated with each population and create a file that we can use to specify calculations of $F$ for only those sets of samples. This way, the expected heterozygosities will be calculated on a per-population basis.  
```{r makeKeepLists}
#IDs for individuals in the VCF
vcfHead <- system("grep ^#CHROM PopGen6.vcf | tr '\t' '\n' | grep ^UGA", 
                  intern = TRUE)

#Build a popmap file to map individual samples back to their population
popmap <- data.frame(vcfID = vcfHead, ugaID = NA)
popmap$region <- popmap$year <- NA
popmap$survivor <- popmap$finalClone <- NA
for(i in 1:nrow(popmap)) {
  popmap$region[i] <- samps$region[which(samps$vcfID== popmap$vcfID[i])]
  popmap$ugaID[i] <- samps$ugaID[which(samps$vcfID== popmap$vcfID[i])]
  popmap$year[i] <- samps$year[which(samps$vcfID== popmap$vcfID[i])]
  popmap$finalClone[i] <- samps$finalClone[which(samps$vcfID== popmap$vcfID[i])]
  #How many regions for this clone?
  nReg <- length(unique(samps$region[which(samps$finalClone==popmap$finalClone[i])]))
  if(nReg > 1) {
    popmap$survivor[i] <- TRUE
  } else {
    popmap$survivor[i] <- FALSE
  }
}
nrow(popmap)

#Subsets for each population
CGA12pop <- popmap[which(popmap$region== "CGA-H" | popmap$survivor== TRUE),]
CGA23pop <- popmap[which(popmap$region== "CGA" | popmap$survivor== TRUE),]
NGApop <- popmap[popmap$region== "NGA",]
SGApop <- popmap[popmap$region== "SGA",]

#Write files for population-level F calculations
write.table(CGA12pop$vcfID, file = "CGA12_Fcheck.txt",
            quote = FALSE,row.names = FALSE,col.names = FALSE)
write.table(CGA23pop$vcfID, file = "CGA23_Fcheck.txt",
            quote = FALSE,row.names = FALSE,col.names = FALSE)
write.table(NGApop$vcfID, file = "NGA_Fcheck.txt",
            quote = FALSE,row.names = FALSE,col.names = FALSE)
write.table(SGApop$vcfID, file = "SGA_Fcheck.txt",
            quote = FALSE,row.names = FALSE,col.names = FALSE)
```

Then we calculate $F$ for each subset...  
```{r engine="bash", comment = ""}
vcftools --vcf PopGen6.vcf --keep CGA12_Fcheck.txt --out CGA12_Fcheck --het
vcftools --vcf PopGen6.vcf --keep CGA23_Fcheck.txt --out CGA23_Fcheck --het
vcftools --vcf PopGen6.vcf --keep NGA_Fcheck.txt --out NGA_Fcheck --het
vcftools --vcf PopGen6.vcf --keep SGA_Fcheck.txt --out SGA_Fcheck --het
```

Read these files into R and create a list of samples to remove...  
```{r checkF}
#Then read these outputs into R & check range of F
CGA12 <- read.table("CGA12_Fcheck.het", header = TRUE)
range(CGA12$F)
CGA23 <- read.table("CGA23_Fcheck.het", header = TRUE)
range(CGA23$F)
NGA <- read.table("NGA_Fcheck.het", header = TRUE)
range(NGA$F)
SGA <- read.table("SGA_Fcheck.het", header = TRUE)
range(SGA$F)

#Remove samples with F < -0.5
drop.us <- CGA23$INDV[which(CGA23$F < -0.5)]
drop.us 

#Write a file with IDs to remove
write.table(drop.us, file = "lowFfilt.txt", quote=FALSE, row.names=FALSE, col.names=FALSE)
```


Remove the samples from the VCF and check that all loci are still biallelic...  
```{r engine="bash", comment = ""}
vcftools --vcf PopGen6.vcf --remove lowFfilt.txt --out PopGen7 --recode
vcftools --vcf PopGen7.recode.vcf --out PopGen7 --freq
```

Read this output into R and confirm that all loci are biallelic in this sample.  
```{r checkPoly2}
#Read frequency file into R
snpfreq <- read.table("PopGen7.frq", skip = 1, fill=TRUE)
colnames(snpfreq) <- c("CHROM","POS","N_ALLELES","N_CHR", "A1","A2")

#Which loci are reported as biallelic?
biallelic <- snpfreq[which(snpfreq$N_ALLELES == 2),]
nrow(biallelic)

#Are any of these fixed (freq of A1 or A2 = 1)?
fixed.sites <- c(grep(":1", biallelic$A1), grep(":1", biallelic$A2))
fixed.sites

#Fixed sites should also be removed
biallelic[fixed.sites,]

#List of sites that are indeed biallelic in the final dataset
confirmed.biallelic <- biallelic[-fixed.sites,c(1,2)]
nrow(confirmed.biallelic)

#Write a file for vcftools to filter to these sites...
write.table(confirmed.biallelic, file = "confirmed.biallelic.txt",
            quote = FALSE, row.names = FALSE, col.names = FALSE)
```

Now back to bash one last time to write the final output...  
```{r engine="bash", comment = ""}
vcftools --vcf PopGen7.recode.vcf --out PopGen \
          --positions confirmed.biallelic.txt \
          --remove-filtered-all --recode
          
mv PopGen.recode.vcf PopGen.vcf
```

```{r echo=FALSE}
snpct <- as.numeric(system("grep -v ^# PopGen.vcf | wc -l", intern=TRUE))
```

## Check the final VCF
A few summaries of the dataset. At this point we have `r snpct` variable sites in the `.vcf` file.  
```{r engine="bash", comment = ""}
#Output Hardy-Weinberg Equilibrium tests
vcftools --vcf PopGen.vcf --out PopGen --hardy
```

How many sites are out of HWE?  
```{r checkHWE}
hwe.res <- read.table("PopGen.hwe", header = TRUE)
range(hwe.res$P_HWE)
hist(hwe.res$P_HWE, main = "Distribution of HWE p-values across SNP loci")
legend("topleft", bty = "n", legend = c(paste("n =", nrow(hwe.res), "SNPs"),
                                        paste("min p-value =", 
                                              round(min(hwe.res$P_HWE),3))))
```

Downstream analyses are presented in `2_diversity`, `3_structure`, and `4_effectiveSize`.  

---



